---
title: "Confirmatory Factor Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

In this tutorial, we are going to use `lavaan` for confirmatory factor analysis. 


## Load the pacakges

```{r, message=FALSE, warning=FALSE}
library(lavaan)
```

<br>

## Example 1: One-factor CFA model

### Import the data

The data for this example is saved in a csv file named "proficiencyraw-female.csv", which is provided as part of the course materials. You can use the `read.table()` function to read in a csv file. There are many different ways for you to supply the file path in the `read.table()` function call. Below I first set the working directory to the folder where my data is saved, and then I simply supply the name of the data file in the function call. 

```{r,echo=F}
mypath <- "/Volumes/GoogleDrive/My Drive/Dpr/EDMS workshops/SEM2019/SEM First Course Mplus 2019/Mplus Examples"
```

```{r}
setwd(mypath)  # change it to the path of your own data folder
data <- read.table("proficiencyraw-female.csv", sep=",", header=F)

data <- data[, 7:11]
# give variable names
colnames(data) <- paste0("rsc",1:5)

```

You can take a look at the data and get the sample statistics.

```{r}

head(data)
str(data)

# sample means
apply(data, 2, mean)

# sample correlations
knitr::kable(cor(data))

```


<br>


---


We now fit a one factor model to the data. 

```{r, echo = F, warnings = F, message = F, out.width='80%'}
library(semPlot)

cfa.Model <- '

# measurement model

RSC =~ rsc1 + rsc2 + rsc3 + rsc4 + rsc5

# mean structure
rsc1~1
rsc2~1
rsc3~1
rsc4~1
rsc5~1

# variances

rsc1 ~~ rsc1
rsc2 ~~ rsc2
rsc3 ~~ rsc3
rsc4 ~~ rsc4
rsc5 ~~ rsc5

'
cfa.Fit <- sem(cfa.Model, data = data)
semPaths(cfa.Fit, title = FALSE, curvePivot = TRUE, intercepts = F, optimizeLatRes = T)

```

### Write the model syntax

We use the model syntax to tell `lavaan` what model we want to fit. Different from our previous MLR and MVPA examples, we now have latent variables that are not observed in the data. For `lavaan` to recognize the latent variables, we must define each of them using the special operator "`=~`", which is followed by the indicators of the latent variable being defined. You can read the "`=~`" operator as "is measured by".

In our example, we have to define one latent variable: RSC (F1). By default, `lavaan` will fix the loading of the first indicator to 1 (similar with Mplus).


```{r}

cfa.Model <- '

# measurement model

RSC =~ rsc1 + rsc2 + rsc3 + rsc4 + rsc5

# mean structure
rsc1~1
rsc2~1
rsc3~1
rsc4~1
rsc5~1

# variances

rsc1 ~~ rsc1
rsc2 ~~ rsc2
rsc3 ~~ rsc3
rsc4 ~~ rsc4
rsc5 ~~ rsc5

'

```

<br>

### Fit the model to the data

We can still use the `sem()` function for CFA. You just need to simply supply the model syntax and the data. 

```{r, eval = FALSE}
cfa.Fit <- sem(cfa.Model, data = data)

```

<br>

### Summary the results

Next you can summarize the results using the `summary()`function. To get more info on the model fit measures, you can use the optional argument `fit.measures = TRUE`. To take a look at the standardzied parameter estimates, you can use the optional argument `standardized = TRUE`. Unstandardized solutions are displayed in the column labled "Estimate". The standardized solutions are listed in the column labeled "Std.all".

```{r}
summary(cfa.Fit, fit.measures = T, standardized = T)
```


---

## Example 2: Three-factor CFA model


### Import the data

```{r, echo = F, warnings = F, message = F, out.width='80%'}

covmat <- '
0.870           
0.345 0.765          
0.270 0.333 1.775         
0.312 0.279 0.446 0.781        
-0.210 -0.301 -0.386 -0.364 2.182       
-0.223 -0.200 -0.371 -0.451 0.936 2.178      
-0.235 -0.234 -0.369 -0.347 0.876 1.520 2.283     
-0.207 -0.236 -0.220 -0.240 0.704 0.595 0.601 1.255    
-0.065 -0.095 -0.325 -0.309 0.625 0.749 0.850 0.462 2.752   
-0.078 -0.085 -0.253 -0.318 0.512 0.640 0.741 0.379 1.915 2.472  
-0.136 -0.226 -0.518 -0.376 0.854 0.746 0.930 0.512 1.709 1.545 2.924 
-0.106 -0.253 -0.295 -0.415 1.016 0.869 0.867 0.589 1.198 1.097 1.349 2.747
'

cfa.cov <- getCov(covmat,names=c("i4", "i7", "i9", "i17", "i5", "i10", "i11", "i15", "i1", "i2", "i3", "i6"))

cfa.model <- '
EXHAUST =~ 1*i1 + i2 + i3 + i6
DEPERSON =~ 1*i5 + i10 + i11 + i15
ACCOMP =~ i4 + i7 + i9 + i17

EXHAUST ~~ DEPERSON + ACCOMP
DEPERSON ~~ ACCOMP

'

cfa.fit <- sem(cfa.model, sample.cov = cfa.cov, sample.nobs = 580)

semPaths(cfa.fit, title = FALSE, curvePivot = TRUE, intercepts = F, optimizeLatRes = T)

```


```{r}
covmat <- '
0.870           
0.345 0.765          
0.270 0.333 1.775         
0.312 0.279 0.446 0.781        
-0.210 -0.301 -0.386 -0.364 2.182       
-0.223 -0.200 -0.371 -0.451 0.936 2.178      
-0.235 -0.234 -0.369 -0.347 0.876 1.520 2.283     
-0.207 -0.236 -0.220 -0.240 0.704 0.595 0.601 1.255    
-0.065 -0.095 -0.325 -0.309 0.625 0.749 0.850 0.462 2.752   
-0.078 -0.085 -0.253 -0.318 0.512 0.640 0.741 0.379 1.915 2.472  
-0.136 -0.226 -0.518 -0.376 0.854 0.746 0.930 0.512 1.709 1.545 2.924 
-0.106 -0.253 -0.295 -0.415 1.016 0.869 0.867 0.589 1.198 1.097 1.349 2.747
'

cfa.cov <- getCov(covmat,names=c("i4", "i7", "i9", "i17", "i5", "i10", "i11", "i15", "i1", "i2", "i3", "i6"))

```

### Write the model syntax

```{r}
cfa.model <- '
EXHAUST =~ 1*i1 + i2 + i3 + i6
DEPERSON =~ 1*i5 + i10 + i11 + i15
ACCOMP =~ i4 + i7 + i9 + i17

EXHAUST ~~ DEPERSON + ACCOMP
DEPERSON ~~ ACCOMP

'
```


### Fit the model to the data and check the results

```{r}
cfa.fit <- sem(cfa.model, sample.cov = cfa.cov, sample.nobs = 580)
summary(cfa.fit, fit.measures=T, standardized = T)

```


### Check modification indicies

We could check the modification incides and print out the ones exceeding the critical value 3.84. 

```{r}
mod.fit <- modindices(cfa.fit)
with(mod.fit, subset(mod.fit[order(mi, decreasing = T), ], subset = mi >= 3.84))

```


### Modify the model specification 

```{r}
cfa.model2 <- '
EXHAUST =~ 1*i1 + i2 + i3 + i6
DEPERSON =~ 1*i5 + i10 + i11 + i15
ACCOMP =~ i4 + i7 + i9 + i17

EXHAUST ~~ DEPERSON + ACCOMP
DEPERSON ~~ ACCOMP

i1 ~~ i2
i10 ~~ i11

'

cfa.fit2 <- sem(cfa.model2, sample.cov = cfa.cov, sample.nobs = 580)
summary(cfa.fit2, fit.measures=T, standardized = T)

```


---

## Example 3 Second-order self-concept model

In this example, we fit a second-order factor model to the data. 

```{r}
setwd(mypath)  # change it to the path of your own data folder
data <- read.table("proficiencyraw-female.csv", sep=",", header=F)

# give variable names
colnames(data) <- c(paste0("goals",1:6), 
                    paste0("rsc",1:5), 
                    paste0("hsc",1:5), 
                    paste0("msc",1:5), 
                    paste0("ssc",1:5), 
                    "SATvoc", 
                    "SATcomp", 
                    "SATlang", 
                    "SATmath", 
                    "SATprob", 
                    "SATproc")

```

```{r, echo = F, warnings = F, message = F, out.width='80%'}

cfa.Model.2='
RSC=~rsc1+rsc2+rsc3
HSC=~hsc1+hsc2+hsc3
MSC=~msc1+msc2+msc3
SSC=~ssc1+ssc2+ssc3

GENERAL=~NA*RSC+HSC+MSC+SSC
GENERAL~~1*GENERAL

rsc1~1
rsc2~1
rsc3~1
hsc1~1
hsc2~1
hsc3~1
msc1~1
msc2~1
msc3~1
ssc1~1
ssc2~1
ssc3~1


rsc1~~hsc1+msc1+ssc1
rsc2~~hsc2+msc2+ssc2
rsc3~~hsc3+msc3+ssc3

hsc1~~msc1+ssc1
hsc2~~msc2+ssc2
hsc3~~msc3+ssc3

msc1~~ssc1
msc2~~ssc2
msc3~~ssc3

'

cfa.Fit.2=sem(cfa.Model.2,data=data)
semPaths(cfa.Fit.2, title = FALSE, curvePivot = TRUE, intercepts = F)

```

### Write the model syntax

We first define the first-order latent factors, as we did in the previous example. The second-order latent factor "GENERAL" is also defined using the "`=~`" operator. Its indicators are the first-order latent factors. We can free the factor loading by multiplying the indicator with `NA`, such that lavaan knows that the value of this parameter is not fixed and needs to be estimated. We then fix the variance of the second-order latent factor to be 1, by multiplying it with 1 in the variance formula. 


```{r}
cfa.Model.2 <- '

# measurement model 

RSC =~ rsc1 + rsc2 + rsc3
HSC =~ hsc1 + hsc2 + hsc3
MSC =~ msc1 + msc2 + msc3
SSC =~ ssc1 + ssc2 + ssc3

# second-order latent factor

GENERAL =~ NA*RSC + HSC + MSC + SSC
GENERAL ~~ 1*GENERAL

# mean structure
rsc1~1
rsc2~1
rsc3~1
hsc1~1
hsc2~1
hsc3~1
msc1~1
msc2~1
msc3~1
ssc1~1
ssc2~1
ssc3~1


# residual covariances 

rsc1~~hsc1+msc1+ssc1
rsc2~~hsc2+msc2+ssc2
rsc3~~hsc3+msc3+ssc3

hsc1~~msc1+ssc1
hsc2~~msc2+ssc2
hsc3~~msc3+ssc3

msc1~~ssc1
msc2~~ssc2
msc3~~ssc3

'

```

<br>

### Fit the model to the data

We can still use the `sem()` function for second-order CFA. You just need to simply supply the corresponding model syntax and the data. 

```{r, eval = FALSE}
cfa.Fit.2 <- sem(cfa.Model.2,data=data)
```

<br>

### Summary the results

Next you can summarize the results using the `summary()`function. To get more info on the model fit measures, you can use the optional argument `fit.measures = TRUE`. To take a look at the standardzied parameter estimates, you can use the optional argument `standardized = TRUE`. Unstandardized solutions are displayed in the column labled "Estimate". The standardized solutions are listed in the column labeled "Std.all".

```{r}
summary(cfa.Fit.2, fit.measures = T, standardized = T)
```

---


## Example 4: Residualized factor model

In this example, we fit a residualized factor model to the self-concept data. 

```{r, echo = F, warnings = F, message = F, out.width='80%'}

cfa.Model.3='
RSC=~rsc1+rsc2+rsc3
HSC=~hsc1+hsc2+hsc3
MSC=~msc1+msc2+msc3
SSC=~ssc1+ssc2+ssc3

GENERAL=~NA*rsc1+rsc2+rsc3+hsc1+hsc2+hsc3+msc1+msc2+msc3+ssc1+ssc2+ssc3
GENERAL~~1*GENERAL

GENERAL~~0*RSC+0*HSC+0*MSC+0*SSC
RSC~~0*HSC+0*MSC+0*SSC
HSC~~0*MSC+0*SSC
MSC~~0*SSC

rsc1~1
rsc2~1
rsc3~1
hsc1~1
hsc2~1
hsc3~1
msc1~1
msc2~1
msc3~1
ssc1~1
ssc2~1
ssc3~1


rsc1~~hsc1+msc1+ssc1
rsc2~~hsc2+msc2+ssc2
rsc3~~hsc3+msc3+ssc3

hsc1~~msc1+ssc1
hsc2~~msc2+ssc2
hsc3~~msc3+ssc3

msc1~~ssc1
msc2~~ssc2
msc3~~ssc3

'

cfa.Fit.3 <- sem(cfa.Model.3,data=data)
semPaths(cfa.Fit.3, title = FALSE, curvePivot = TRUE, intercepts = F, optimizeLatRes = T)

```

### Write the model syntax

The residualized latent factor "GENERAL" is defined by all the measured variables. We can free the factor loading by multiplying the first indicator with `NA`. `lavaan` thus is told that the value of this parameter is not known and needs to be estimated. We then fix the variance of the residualized latent factor to be 1, by multiplying it with 1 in the variance formula. 

We also fix the covariances among the latent factors to be 0. In order to do that, we multiply the variables by 0 in each of the variance formulas.


```{r}
cfa.Model.3 <- '

# measurement model

RSC =~ rsc1 + rsc2 + rsc3
HSC =~ hsc1 + hsc2 + hsc3
MSC =~ msc1 + msc2 + msc3
SSC =~ ssc1 + ssc2 + ssc3


# residualized latent factor 

GENERAL =~ NA*rsc1 + rsc2 + rsc3 + hsc1 + hsc2 + hsc3 + msc1 + msc2 + msc3 + ssc1 + ssc2 + ssc3
GENERAL ~~ 1*GENERAL


# fix the covariances to be 0

GENERAL ~~ 0*RSC + 0*HSC + 0*MSC + 0*SSC
RSC ~~ 0*HSC + 0*MSC + 0*SSC
HSC ~~ 0*MSC+0*SSC
MSC ~~ 0*SSC


# mean structure

rsc1~1
rsc2~1
rsc3~1
hsc1~1
hsc2~1
hsc3~1
msc1~1
msc2~1
msc3~1
ssc1~1
ssc2~1
ssc3~1


# residual covariances 

rsc1 ~~ hsc1 + msc1 + ssc1
rsc2 ~~ hsc2 + msc2 + ssc2
rsc3 ~~ hsc3 + msc3 + ssc3

hsc1 ~~ msc1+ssc1
hsc2 ~~ msc2+ssc2
hsc3 ~~ msc3+ssc3

msc1 ~~ ssc1
msc2 ~~ ssc2
msc3 ~~ ssc3

'

```

<br>

### Fit the model to the data

```{r, eval = FALSE}
cfa.Fit.3 <- sem(cfa.Model.3, data = data)
```

<br>

### Summary the results

```{r}
summary(cfa.Fit.3, fit.measures = T, standardized = T)
```


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>

